---
title: "Data Requirements and Validation"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Data Requirements and Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(artfishr)
```

## 1. Introduction

The R package `artfishr` implements tools for producing catch and effort estimates from sample-based survey data from small-scales fisheries data following the *FAO ARTFISH extrapolation methodology*.  
To ensure consistency and interoperability, input datasets must comply with standardized formats and validation rules.

This vignette describes:

- the required **data formats** used in `artfishr`,
- the **sample data** and format specifications provided with the package,
- how to **generate empty templates** for each input data,
- and how to **validate** input data.

Four core datasets are supported:

1. **Active vessels**: Number of active vessels per stratum (vessels that can potentially go fishing) 
2. **Effort** (two variants depending on data collection method): fisher interview or boat counting  
3. **Active days**: number of days in the month with a fishing activity
4. **Landings**: landings survey observation data

---

## 2. Overview of ARTFISH data formats

Each dataset type follows a standardized structure defined by a JSON specification file.  
These schemas describe the column names, units, and validation rules.

| Dataset | Specification | Sample data |
|----------|----------------|--------------|
| Active vessels | `inst/extdata/format_specs/artfish_A_active_vessels.json` | `inst/extdata/samples/active_vessels.csv` |
| Effort (fisher interviews) | `inst/extdata/format_specs/artfish_B1_effort.json` | `inst/extdata/samples/effort.csv` |
| Effort (boat counting) | `inst/extdata/format_specs/artfish_B2_effort.json` | `(To add)` |
| Active days | `inst/extdata/format_specs/artfish_C_active_days.json` | `inst/extdata/samples/active_days.csv` |
| Landings | `inst/extdata/format_specs/artfish_D_landings.json` | `inst/extdata/samples/landings.csv` |

Each example file can be accessed directly via `system.file()`:

```{r, eval=FALSE}
system.file("extdata/samples/active_vessels.csv", package = "artfishr")
```

Preview the structure:

```{r, eval=FALSE}
readr::read_csv(system.file("extdata/samples/active_vessels.csv", package = "artfishr")) |>  head()
```

The specification file describes each field and the required data structure.

---

## 3. Identify your sampling strategy

(Text to improve including effort type and cases where active_days is need or not)

Effort data can be collected under two distinct monitoring approaches:

1. **Fisher interviews (`effort_source = "fisher_interview"`)**  
   - Data obtained directly from fishers with questionnaires    
   - The template `effort.csv` is required (number of days fished within the period of reference)
   - In this case, the active days survey is not necessary. Total number of days in the month is used by default.
  

2. **Boat counting (`effort_source = "boat_counting"`)**  
   - Data collected via registers or direct observation
   - Template **y (to complete)** is required (number of boats out in comparison of total potentially active boat)
   - The number of active days in the month must be provided. It can come from the active days survey or from estimation based on fisheries knowledge.

Both formats are supported by `artfishr` and can be validated using the same validation interface.

---

## 4. Creating empty data templates

`artfishr` allows users to programmatically generate empty templates for each dataset, based on their JSON specification.  

The package provides a generic method `create_artfish_template()` to generate templates using the dataset names listed above.

```{r, eval=FALSE}
create_artfish_template("artfish_A_active_vessels")
```
You can also export the template for use with the argument `save_as`:

```{r, eval=FALSE}
create_artfish_template("artfish_A_active_vessels",save_as="active_vessels.csv")
```

Or use dedicated functions by dataset : 

```{r, eval=FALSE}
create_active_vessels_template()
create_effort_template(effort_source ="boat_counting") # or "fisher_interview"
create_landings_template(save_as="landings.csv")
create_active_days_template()
```

---

## 5. Validating input datasets

`artfishr` includes validation utilities to ensure all datasets comply with the expected structure and logical rules.

A **global validator** is provided:

```{r, eval=FALSE}
validate_input_datasets(
  active_vessels,
  effort,
  effort_source = c("fisher_interview", "boat_counting"),
  active_days = NULL,
  landings
)
```

This function:
- checks required fields for each dataset and flags missing or unexpected columns,
- checks that field values satisfy the validation rules,
- and ensures consistency between them.

Example:

```{r, eval=FALSE}
# Load sample datasets
active_vessels <- readr::read_csv(system.file("extdata/samples/active_vessels.csv", package = "artfishr"))
effort <- readr::read_csv(system.file("extdata/samples/effort.csv", package = "artfishr"))
landings <- readr::read_csv(system.file("extdata/samples/landings.csv", package = "artfishr"))

# Validate all datasets
validate_input_datasets(
  active_vessels = active_vessels,
  effort = effort,
  effort_source = "fisher_interview",
  landings = landings
)
```

The function returns a structured report with validation results and messages.

`artfishr` also includes **specific validators** to allow running checks per dataset type.

```{r, eval=FALSE}
# Load sample datasets
active_vessels <- readr::read_csv(system.file("extdata/samples/active_vessels.csv", package = "artfishr"))
effort <- readr::read_csv(system.file("extdata/samples/effort.csv", package = "artfishr"))
landings <- readr::read_csv(system.file("extdata/samples/landings.csv", package = "artfishr"))
active_days <- readr::read_csv(system.file("extdata/samples/active_days.csv", package = "artfishr"))

validate_active_vessels_template(active_vessels)
validate_effort_template(effort,effort_source ="boat_counting") # or "fisher_interview"
validate_landings_template(landings)
validate_active_days_template(active_days)
```

---

## 6. Summary and recommendations

- Use the provided JSON specs and templates to ensure data consistency.  
- Always validate your datasets before running analyses.  
- Choose the appropriate *effort survey format* according to your monitoring method.  
- Keep datasets synchronized on mandatory fields (e.g., year, fishing unit, landing site).  
- Extend or customize validation rules where necessary for national contexts.

---

## 7. References

de Graaf, G. J., Nunoo, F., Ofori-Danson, P., Wiafe, G., Lamptey, E., & Bannerman, P. (2015). *International training course in fisheries statistics and data collection*. FAO Fisheries and Aquaculture Circular No. 1091. Rome: Food and Agriculture Organization of the United Nations. 134 pp. Available at: <https://www.fao.org/3/i3639e/i3639e.pdf>

Stamatopoulos, C. (2002). *Sample based fishery surveys: A technical handbook*. FAO Fisheries Technical Paper No. 425. Rome: Food and Agriculture Organization of the United Nations. 132 pp. Available at: <https://www.fao.org/3/y2790e/y2790e.pdf>
