---
title: Introduction to artfishr
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to artfishr}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(artfishr)
```

This vignette shows how to use [artfishr](https://github.com/fdiwg/artfishr).

## Installation

The R [artfishr](https://github.com/fdiwg/artfishr) can be installed from Github. 

As prequirements, packages `remotes` and `vrule` should be installed.

* Package `remotes` can be installed from CRAN using:

```{r, eval=FALSE}
install.packages("remotes", repos = "https://cloud.r-project.org")
```

* Package [vrule](https://github.com/fdiwg/vrule) should be installed from GitHub using `remotes`:

```{r, eval=FALSE}
remotes::install_github("fdiwg/vrule")
```

Once packages `remotes` and `vrule`, R package `artfishr` can be installed using:

```{r, eval=FALSE}
remotes::install_github("fdiwg/artfishr")
```

Once installed, `artfishr` can be loaded using `library(artfishr)` or `require(artfishr)``

## Data requirements

See the full guide: `vignette("02_data-requirements-validation")`

## How to run Artfish with `artfishr`

Full tutorial: `vignette("03_artfishr-workflow")`

## The Artfish methodology - summary
See: `vignette("04_artfish-methodology")`

## Exemple of use

```{r, eval=FALSE}
# Load sample datasets
active_vessels <- read.csv(system.file("extdata/samples/active_vessels.csv", package = "artfishr"))
effort <- read.csv(system.file("extdata/samples/effort.csv", package = "artfishr"))
landings <- read.csv(system.file("extdata/samples/landings.csv", package = "artfishr"))
active_days <- read.csv(system.file("extdata/samples/active_days.csv", package = "artfishr"))

# Validate data 
validate_input_datasets(
  active_vessels = active_vessels,
  effort = effort,
  effort_source = "fisher_interview",
  landings = landings,
  active_days = active_days
)

#Run artfish
report <- artfishr::compute_report(
  active_vessels = active_vessels,
  effort = effort,
  effort_source = "fisher_interview",
  active_days = active_days,
  active_vessels_strategy = "closest",
  landings = landings,
  minor_strata = "minor_stratum"
)

#See result
head(report)

```