---
title: Introduction to artfishr
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to artfishr}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This vignette shows how to use [artfishr](https://github.com/fdiwg/artfishr).

## Installation

The R [artfishr](https://github.com/fdiwg/artfishr) can be installed from Github. 

As prequirements, packages `remotes` and `vrule` should be installed.

* Package `remotes` can be installed from CRAN using:

```{r}
install.packages("remotes", repos = "https://cloud.r-project.org")
```

* Package [vrule](https://github.com/fdiwg/vrule) should be installed from GitHub using `remotes`:

```{r}
remotes::install_github("fdiwg/vrule")
```

Once packages `remotes` and `vrule`, R package `artfishr` can be installed using:

```{r}
remotes::install_github("fdiwg/artfishr")
```

Once installed, `artfishr` can be loaded using `library(artfishr)` or `require(artfishr)``

## Data requirements

TODO

## The Artfish methodology - summary

TODO

## How to run Artfish with `artfishr`

TODO